---
title: "Assignment 3: Data Exploration"
author: "KEY"
date: "Spring 2026"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  [NEW] Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to Canvas.

8.  Initial here to acknowledge that you did not use AI in completing this assignment, except where expressly allowed: ___

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks in your code chunks.

**TIP**: If your code fails to knit, check:
 * That no `install.packages()` or `View()` commands exist in your code. 
 * That you are not displaying the entire contents of a large dataframe in your code.

---

## Set up your R session

1.  Load necessary packages (tidyverse, here), check your current working directory and import two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 

**Be sure to**:
* Use the `here()` package in specifying the paths to your datasets
* Include the appropriate subcommand to read in character based columns as factors


```{r,  message = FALSE}
#Load packages
library(tidyverse); library(here)

#Check working directory
here()

#Import Data
Neonics <- read.csv(
  file=here('Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv'),
  stringsAsFactors = TRUE
)

Litter <- read.csv(
  file=here('Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv'),
  stringsAsFactors = TRUE
)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. (AI is allowed here, but put answers in your own words.)

> Answer: 
This dataset may shed light on the costs of neonicitoid application; specifically, are these pesticides also affecting pollinators, which are key to the success of many crops.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. (AI is allowed here, but put answers in your own words.)

> Answer:
Litter and woody debris are important components of forest ecosystems. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter and fine woody debris are collected from elevated and ground traps, respectively.
 2. Mass data for each collection event are measured separately for the 8 functional groups.
 3. Litter and fine woody debris sampling is executed at terrestrial NEON sites that contain woody vegetation >2m tall. 


## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
#Report dimensions
dim(Neonics)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied.  [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}
#Report the summary of the Effect column
summary(Neonics$Effect) |> sort(decreasing = T)
```

► Question: Which two effects stand out as the most studied? Can you guess why these effects might specifically be of interest?
> Answer:
Population and Mortality. These are the easiest to define and most important. 

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name).[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
#Show the top 6 most studied species
summary(Neonics$Species.Common.Name,maxsum = 6) |> sort(decreasing = T)

```

► Question: What do these species have in common? Why might they be of interest over other insects? 
> Answer:
Most are bees, which are important to pollination. 

8.  The `Conc.1..Author` column, which lists the concentration of the neonicitoid dose, should include numeric values. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}
#Show the class of the `Conc.1..Author` column
class(Neonics$Conc.1..Author.)

```

> Answer:
It's a factor because the column includes some non-numeric values. 


## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
#Create a frequency polyline plot of publications by year
ggplot(Neonics,aes(x=Publication.Year)) + 
  geom_freqpoly()

```


10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
#Create a frequency polyline plot of publications by year, colored by location
ggplot(Neonics,aes(x=Publication.Year, color=Test.Location)) + 
  geom_freqpoly()

```


►Interpret this graph. What are the most common test locations, and do they differ over time?
> Answer:
The lab is usually the most common location, particularly in most recent studies. Field natural is the second most common, and in some earlier years is the most common. 

11. Create a bar graph of Endpoint counts. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(Neonics, aes(x=Endpoint)) + 
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


► What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix (p.721) for more information. 
> Answer: LOEL (Lowest-observable-effect-level) & NOEL (No-observable-effect-level)

---

## Explore your data (Litter)

12. Determine the class of `collectDate`. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
#Report the class
class(Litter$collectDate)

#Set to date 
Litter$collectDate <- ymd(Litter$collectDate)

#Report the class
class(Litter$collectDate)

```


13. Using the `unique` function, list the different `plotIDs` sampled at Niwot Ridge. 

```{r}
#List unique plotID values
unique(Litter$plotID)
summary(Litter$plotID)
```

► How is the information obtained from `unique` different from that obtained from `summary`?
> Answer:
`unique()` reports only the labels; `summary()` also reports the counts of each level

14. Create a bar graph of `functionalGroup` counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
#Bar plot of functional group counts
ggplot(Litter, aes(x=functionalGroup)) +
  geom_bar()
```


15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of `dryMass` by `functionalGroup`.

```{r}
#Box plot of functional group counts
ggplot(Litter, aes(x=dryMass, y=functionalGroup)) +
  geom_boxplot()

#Violin plot of functional group counts
ggplot(Litter, aes(x=dryMass, y=functionalGroup)) +
  geom_violin()
```

► Why is the boxplot a more effective visualization option than the violin plot in this case?
> Answer: Very narrow, long distributions are hidden in violin plots


► What type(s) of litter tend to have the highest biomass at these sites?
> Answer:
Needles.